# Statistical Inference and Causality

## Introduction
Statistical inference is the process of using data analysis to deduce properties of an underlying distribution of probability. In other words: statistical inference is the procedure through which we try to make inferences about a population based on characteristics of this population that have been captured in a sample. This includes uncovering the association between variables as well as establishing causal relationships. However, correlation does not imply causation and identifying causal relationships from observational data is not trivial.

> "Correlation does not imply causation" (Any statistician).

This week we will dive a little deeper into statistical inference and how to establish causal relationships by taking a small dive into the field of study of [econometrics](https://en.wikipedia.org/wiki/Econometrics#:~:text=Econometrics%20is%20the%20application%20of,by%20appropriate%20methods%20of%20inference%22.). This week is structured by **4** short lecture videos, reading material, a tutorial in R with a 'hands-on' application of the techniques covered in the lecture videos, and a seminar on Tuesday. The short lecture videos this week are provided by an absolute expert in the field of econometrics: [Dr Idahosa](https://www.uj.ac.za/contact/Pages/Love-Idahosa.aspx) from the [University of Johannesburg](https://www.uj.ac.za/).

Let's get to it!

#### Video: Introduction W01 {-}
```{r 01-short-lecture-welcome, warnings=FALSE, message=FALSE, echo=FALSE}
library(vembedr)
embed_msstream('') %>% use_align('left')
```
[Lecture slides] [[Watch on MS stream]]()

```{r 01-settings, warnings=FALSE, message=FALSE, echo=FALSE}
# settings
options(max.print=30)
```

### Reading list {#reading-list-w01}
Please find the reading list for this week below.

#### Core reading {-}
- Bzdok, D., Altman, and M. Krzywinski. 2018. Statistics versus machine learning. *Nature Methods* 15: 233-234. [[Link]](https://www.nature.com/articles/nmeth.4642)
- Idahosa, L. O., Marwa, N., and J. Akotey. 2017. Energy (electricity) consumption in South African hotels: A panel data analysis. *Energy and Buildings* 156: 207-217. [[Link]](https://ucl-new-primo.hosted.exlibrisgroup.com/permalink/f/1klfcc3/TN_cdi_gale_infotracacademiconefile_A523007644)
- Stock, J. and M. Watson. 2019. **Chapter 1**: *Economic Questions and Data*. In: Stock, J. and M. Watson. Introduction to Econometrics, pp.43-54. Harlow: Pearson Education Ltd. [[Link]](https://ucl-new-primo.hosted.exlibrisgroup.com/permalink/f/1klfcc3/TN_cdi_proquest_ebookcentral_EBC6399072)
- Stock, J. and M. Watson. 2019. **Chapter 10**: *Regression with Panel Data*. In: Stock, J. and M. Watson. Introduction to Econometrics, pp.362-382. Harlow: Pearson Education Ltd. [[Link]](https://ucl-new-primo.hosted.exlibrisgroup.com/permalink/f/1klfcc3/TN_cdi_proquest_ebookcentral_EBC6399072)

#### Supplementary reading {-}
- Idahosa, L. O. and J. T. van Dijk. 2016. South Africa: Freedom for whom? Inequality, unemployment and the elderly. *Development* 58(1): 96-102. [[Link]](https://ucl-new-primo.hosted.exlibrisgroup.com/permalink/f/1klfcc3/TN_cdi_proquest_journals_1758956430)

### Technical Help session
Every Thursday between 12h00-13h00 you can join the **Technical Help** session on Microsoft Teams. The session will be hosted by [Alfie](https://www.ucl.ac.uk/geospatial-analytics/people/alfie-long). He will be there for the whole hour to answer any question you have live in the meeting or any questions you have formulated beforehand. If you cannot make the meeting, feel free to post the issue you are having in the Technical Help channel on the GEOG0125 Team so that Alfie will can help to find a solution. 

## Statistical inference
Where during the remainder of this module we will predominantly focus on different [machine learning](https://en.wikipedia.org/wiki/Machine_learning#:~:text=Machine%20learning%20(ML)%20is%20the,a%20subset%20of%20artificial%20intelligence.) methods and techniques) and the data scientist's toolbox, this week we will focus on statistical inference. Although there is considerable overlap sometimes even involving the same methods, the major difference between machine learning and statistics is their purpose. Machine learning models are designed to make **accurate predictions**. Statistical models are designed for inference about the **relationships between variables**. 

>"Statistics draws population inferences from a sample, and machine learning finds generalizable
predictive patterns" (Bzdok *et al.* 2018).

Before we move onto the more complicated material, let's have a quick refresher on linear regression models. Linear regression models offer a very **interpretable** way to model the association between variables. A linear regression model is used to find the line that minimises the mean squared error accross all data to establish the relationship between a response variable one or more explanatory (independent) variables. The case of one explanatory variable is called a **simple linear regression**; if more explanatory variables are used it is called **multiple regression**.

#### Video: Linear regression models {-}
```{r 02-video-linear-models, warnings=FALSE, message=FALSE, echo=FALSE}
library(vembedr)
embed_msstream('') %>% use_align('left')
```
[[Lecture slides]]() [[Watch on MS stream]]()

## Causality
### What is causality?
It should now be clear that regression is a statistical technique with the purpose to predict a target variable $y$ according to some other variable $x$ or variables $x_{1}$, $x_{2}$, etc. Regression, however, cares about correlation: what happens to $Y$ when $X=x$. However, correlation does not imply causation. A great example is the relationships between [chocolate consumption and the probability of winning a Nobel price](https://www.reuters.com/article/us-eat-chocolate-win-the-nobel-prize/eat-chocolate-win-the-nobel-prize-idUSBRE8991MS20121010): The higher a country’s chocolate consumption, the more Nobel laureates it spawns per capita. There is a clear correlation, but there should not exist an actual causal relationship between these two variables (Well, some research suggest it still remains [unclear whether the correlation is spurious](https://doi.org/10.1016/j.ssaho.2020.100082) or an indication for hidden variables ...)

The golden standard to establish a causal relationship is to set-up and execute a [randomised control trial](https://adc.bmj.com/content/90/8/840), think of the many large-scale randomised control trials that are currently taking place to test the safety and effectiveness of various candidate coronavirus vaccines. However, it is not always possible to set up a randomised control trial. Sometimes this has to do with the nature of the relationship being investigated (e.g. establishing the effects of policy changes), but there could also be financial and ethical constraints. As an alternative, one could try to identify causal relationship from observational data. This is known as causal inference and most research in econometrics is concerned with retrieving valid estimates using different regression methods.

<div class="note">
The distinction between causal and non-causal relationships is crucial and heavily depends on your research questions. If you are trying to classify Google Street view images and predict whether the photo contains an office building or a residential building, you want to create a model that predicts this as good as possible. However, you do not care of what was the cause of the building in the photo to be an office building or a residential building. That being said: almost any question is causal and where statistics is used in almost any field of inquiry, few pay proper attention to understanding causality.
</div>

By now you may wonder, sure, but what then is causality? We can say that $X$ causes $Y$ if we were to change the value of $X$ without changing anything else then as a result $Y$ would also change. A simple example: if you switch on the light switch, your light will go on. The action of flipping the light switch causes your light to go on. This being said: it does not mean that $X$ is necessarily the only thing that causes $Y$ (e.g. the light bulb is burnt out or the light was already on) and perhaps a better way of phrasing it is to say that there is a causal relationship between variables if $X$ changes the probabiblity of $Y$ happening or changing.

### The problem with causality
The problem with establishing a causal relationship is that in many cases you cannot 'switch on' or 'switch off' a characteristic. Let's go through this by thinking whether some $X$ causes $Y$. Our $X$ is coded as 0 or 1, for instance, 0 if a person has not received a coronavirus vaccination and 1 if a person has received a coronavirus vaccination. $Y$ is some numeric value. So how do we check if $X$ causes $Y$? What we would need to do for everyone in our sample is to check what will happen to $Y$ when we make $X=0$ and when we make $X=1$. But this is obviously a problem! You cannot both have $X=0$ and $X=1$: you either got inoculated against the coronavirus or you did not. This means that if $X=1$ you can measure what the value of $Y$ is, but you do not know what the value of $Y$ **would have been** if $X=0$. A solution you may come up with is to compare $Y$ between individuals who have $X=0$ and $X=1$. However, there is another problem: there could be all kinds of reasons on why $Y$ differs between individuals that are not necessarily related to $X$. 

<div class="note">
This section heavily borrows material and explanations from Nick Huntington-Klein's excellent [ECON 305: Economics, Causality, and Analytics](http://www.nickchk.com/econ305.html) module, do have a look if you want to learn more.
</div>

This brings us to econometrics and causal inference: the main goal of causal inference is to make the best possible estimation of what $Y$ would have been if $X$ would have been different, the so-called **counterfactual**. As we cannot always use an experiment in which we can randomly assign $X$ so that we know that on average people with $X=1$ and the same as people with $X=2$, we have to come up with a model to figure out what the counterfactual would do. In the following, we will explore two ways of doing this through so-called fixed-effect models and random effect-models.

### Fixed-effect models
#### Video: Fixed-effect models {-}
```{r 02-video-fixed-effect-models, warnings=FALSE, message=FALSE, echo=FALSE}
embed_msstream('') %>% use_align('left')
```
[[Lecture slides]]() [[Watch on MS stream]]()

### Random-effect models
#### Video: Random-effect models {-}
```{r 02-video-random-effect-models, warnings=FALSE, message=FALSE, echo=FALSE}
embed_msstream('') %>% use_align('left')
```
[[Lecture slides]]() [[Watch on MS stream]]()

## Take home message


## Attributions {#attributions_w02}
This week's content and practical uses content and inspiration from:

- Torres-Reyna, Oscar. 2010. Getting Started in Fixed/Random Effects Models using R. [[Link]](
https://rstudio-pubs-static.s3.amazonaws.com/372492_3e05f38dd3f248e89cdedd317d603b9a.html#45_regression_diagnostics)
- Huntington-Klein, Nick. 2019. ECON 305: Economics, Causality, and Analytics. Lecture 13: Causality. [[Link]](http://www.nickchk.com/econ305.html)
