# Web Visualisation

## Introduction {#intro-w08}
Web mapping or online mapping is the process of using the maps delivered by geographic information systems (GIS) on the Internet. With ever-increasing numbers of large demographic data sets becoming available, many provided as open data, web mapping platforms can be used to facilitate the use and exploration of these datasets. Examples of such web mapping platforms are [DataShine](https://datashine.org.uk/), [CDRC Maps](https://maps.cdrc.ac.uk/), and [CDRC Mapmaker](https://mapmaker.cdrc.ac.uk/).

This week we will dive a little deeper into creating webmaps. We will do this by looking at "the leading open-source JavaScript library for mobile-friendly interactive maps": [Leaflet.js](https://leafletjs.com/). The lecture videos this week are provided by the best mapmaker in UCL: [Oliver O'Brien](https://oobrien.com/). Ollie is a senior research associate at the UCL Department of Geography’s Geospatial Analytics and Computing group. He is particularly interested in investigating and implementing new ways to visualise spatial and socioeconomic data, especially open datasets relating to transport, London and other urban areas. Ollie is also directly reponsible for the creation of [DataShine](https://datashine.org.uk/) and [CDRC Maps](https://maps.cdrc.ac.uk/).

Let's get mapping!

#### Video: Introduction W09 {-}
```{r 09-short-lecture-welcome, warnings=FALSE, message=FALSE, echo=FALSE}
library(vembedr)
embed_msstream('') %>% use_align('left')
```
[Lecture slides] [[Watch on MS stream]]()

```{r 09-settings, warnings=FALSE, message=FALSE, echo=FALSE}
# settings
options(max.print=30)
```

### Reading list {#reading-list-w09}
Because this week's tutorial material is rather extensive and has a strong practical focus, we only have two article as suggested reading this week.

#### Suggested reading {-}
- O'Brien, O. and J. Cheshire. 2015. Interactive mapping for large, open demographic data sets using familiar geographical features. *Journal of Maps* 12(4): 676-683. [[Link]](https://doi.org/10.1080/17445647.2015.1060183)
- Van Dijk, J. T. and P. A. Longley. 2020. Interactive display of surnames distributions in historic and contemporary Great Britain. *Journal of Maps* 16(1): 68-76. [[Link]](https://www.tandfonline.com/doi/full/10.1080/17445647.2020.1746418)

### Technical Help session
Every Thursday between 13h00-14h00 you can join the **Technical Help** session on Microsoft Teams. The session will be hosted by [Alfie](https://www.ucl.ac.uk/geospatial-analytics/people/alfie-long). He will be there for the whole hour to answer any question you have live in the meeting or any questions you have formulated beforehand. If you cannot make the meeting, feel free to post the issue you are having in the Technical Help channel on the GEOG0125 Team so that Alfie can help to find a solution. 

## JavaScript and Leaflet.js
Alongside HTML and CSS, JavaScript (JS) is one of the core technologies of the World Wide Web. JavaScript enables interactive web pages and is an essential part of web applications. The vast majority of websites use it for **client-side** page behavior and all major web browsers have a dedicated JavaScript engine to execute it. Basically you can say that:

- HTML is used to define the content of a web page;
- CSS is used to specify the layout of a web page; and,
- JS to define the behaviour of a web page.

## Attributions {#attributions_w08}
This material was adapted from *JavaScript basics*. [Mozilla Contributors](https://developer.mozilla.org/en-US/docs/Learn/Getting_started_with_the_web/JavaScript_basics/contributors.txt) and is licensed under a Attribution ShareAlike 2.5 Generic license (CC BY-NC-SA 4.0). [[Link]](https://developer.mozilla.org/en-US/docs/Learn/Getting_started_with_the_web/JavaScript_basics)

Futher content and inspiration for this week's material was derived from: the *Interactive Choropleth Map Tutorial* available on https://leafletjs.com/. [[Link]](https://leafletjs.com/examples/choropleth/)

